---
title: "Reproducible Research: Peer Assessment 1"
author: "Cesar"
date: "18/6/2020"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Loading and processing the data

For this analysis we must load and process the raw data. This data was obtained from the internet, and represents the total steps, recorded by a wearable device every 5 minutes of each day in October and November 2012.

```{r, echo=TRUE}
#checks if file exists; if not, downloads zip file
if(!file.exists("./activity.zip")){
        fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileURL,destfile = "./activity.zip", method = "auto")
}

#opens files and loads them to a data frame
steps <- read.csv(unz("./activity.zip","activity.csv"))
```

## 2. What is mean total number of steps taken per day?

Ignoring the missing values of the dataset, we calculate the total steps per day. We make a bar chart from this info. Additionally, we calculate the mean and median from the 5 minute intervals for each day.

```{r, echo=TRUE}
#total steps per day
totalsteps <- aggregate(steps$steps, by = list(steps$date), FUN=sum)
names(totalsteps) <- c("day","total steps")

#Creates a histogram
hist(totalsteps$`total steps`, main = "Frequency of total steps per day", xlab = "Total steps")

#calculates mean and median steps
meansteps <- aggregate(steps$steps, by = list(steps$date), FUN=mean, na.rm=TRUE)
names(meansteps) <- c("day","mean for 5 min")
mediansteps <- aggregate(steps$steps, by = list(steps$date), FUN=median, na.rm=TRUE)
names(mediansteps) <- c("day","median for 5 min")

#Creates a dataframe with all three for convenience
stepsperday <- merge(totalsteps, meansteps, by.x = "day", by.y = "day")
stepsperday <- merge(stepsperday, mediansteps, by.x = "day", by.y = "day")
print(stepsperday)
```

## 3. What is the average daily activity pattern?

Here, I created an analysis of what the average day would look like. It is then ploted into a time series. Furthermore, the data set is analyzed to find the interval with the highest average number of steps.

```{r, echo=TRUE}
#calculates and plots the average day
averageday <- aggregate(steps$steps, by = list(steps$interval), FUN=mean, na.rm=TRUE)
names(averageday) <- c("interval","steps")
plot(averageday, type="l", main = "Average day activity", xlab = "minutes")

#indicates the interval with the highest activity
highestinterval <- averageday[which(averageday$steps == max(averageday$steps)),1]
print(paste("The highest level of activity was, on average, between minute", highestinterval, 
            "and minute", highestinterval+5, "each day."))
```


## 4. Imputing missing values

Given the incompleteness of the dataset, I make an effort to report the ammount of NAs; after this, missing values are replaced with the average for that interval across all days. A new "stepscomplete" dataset is created by this method. Further, to prove the effectiveness of this method, we recalculate the average meassurements from section 2 using the new data.  
The frequency histogram shows that the average is overrepresented in the completed dataset, this is to be expected, given the replacement method. The median is also afected, as it was 0 in most days, but is leveraged up in corrected days (due to most averages being above 0, since at least one day has activity at that interval).

```{r, echo=TRUE}
#counts NA values
numberNA <- sum(is.na(steps$steps))
print(paste("There are", numberNA, "missing values."))

#replaces NA with average values for that interval
stepscomplete <- steps
for (i in 1:17568) {
        if (is.na(stepscomplete[i,]$steps)){
                stepscomplete[i,]$steps <- averageday[which(averageday$interval == stepscomplete[i,]$interval),]$steps
        }
}

#repeats analysis from step 2

totalstepscomplete <- aggregate(stepscomplete$steps, by = list(stepscomplete$date), FUN=sum)
names(totalstepscomplete) <- c("day","total steps")
hist(totalstepscomplete$`total steps`, main = "Frequency of total steps per day", xlab = "Total steps")
meanstepscomplete <- aggregate(stepscomplete$steps, by = list(stepscomplete$date), FUN=mean)
names(meanstepscomplete) <- c("day","mean for 5 min")
medianstepscomplete <- aggregate(stepscomplete$steps, by = list(stepscomplete$date), FUN=median)
names(medianstepscomplete) <- c("day","median for 5 min")
stepsperdaycomplete <- merge(totalstepscomplete, meanstepscomplete, by.x = "day", by.y = "day")
stepsperdaycomplete <- merge(stepsperdaycomplete, medianstepscomplete, by.x = "day", by.y = "day")
print(stepsperdaycomplete)
```

## 5. Are there differences in activity patterns between weekdays and weekends?

We analyze the dataset to answer a final question. A vector was added to the completed dataset inidcating whether a day is a weekday or weekend. The average weekday and weekend day activities were ploted as was done in section 3. We can clearly see more activity on our weekend days.

```{r, echo=TRUE, fig.height= 10}
#added weekday identifier. 
#NOTE: my version of R is in spanish, to replicate this in English change "sábado" to "Satuday" and "domingo" to "Sunday".
stepscomplete$date <- as.Date(stepscomplete$date, format = "%Y-%m-%d")
stepscomplete$weekday <- as.character(1:17568)
for (i in 1:17568) {
        if (weekdays(stepscomplete[i,]$date) == "sábado" || weekdays(stepscomplete[i,]$date) == "domingo"){
                stepscomplete[i,]$weekday <- "weekend"
        } else {
                stepscomplete[i,]$weekday <- "weekday"
        }
}

#average weekday
stepsweekdays <- stepscomplete[which(stepscomplete$weekday == "weekday"),]
averageweekday <- aggregate(stepsweekdays$steps, by = list(stepsweekdays$interval), FUN=mean, na.rm=TRUE)
names(averageweekday) <- c("minutes","steps")

#average weekend day
stepsweekends <- stepscomplete[which(stepscomplete$weekday == "weekend"),]
averageweekends <- aggregate(stepsweekends$steps, by = list(stepsweekends$interval), FUN=mean, na.rm=TRUE)
names(averageweekends) <- c("minutes","steps")

#plots
par(mfrow = c(2,1))
plot(averageweekday, type="l", main = "Average weekday activity")
plot(averageweekends, type="l", main = "Average weekend day activity")
```